Imperfect slope measurements drive overestimation in geometric cone model of waterbody depth

### 2021-11-10 manuscript/manuscript.Rmd
stash tables

As a result of this limited testing, we lack knowledge on both the predictive performance of geometric models, the effect of proxies on depth prediction, and whether depth predictions are more sensitive to measurement errors in the horizontal dimension (i.e. distance to the deepest point of the waterbody) or measurement errors in the vertical dimension (i.e. in-lake slope). Additionally, it is unclear whether model prediction error is related to differences in waterbody type such those with different cross-section shapes (concave versus convex) or those classified as reservoirs versus natural lakes. To address these knowledge gaps, we asked three research questions: (1) How representative is nearshore land slope of in-lake slope; and how representative is the distance to the center of a waterbody compared to the distance to the deepest point of a waterbody? (2) How does the use of proxies for waterbody geometry affect waterbody depth prediction error? (3) How does waterbody cross-section shape (i.e. concave versus convex) and waterbody type (i.e. natural lake vs reservoir) affect depth prediction error? To answer these questions, we extracted maximum depth (hereafter referred to as “observed maximum depth”), in-lake slope, cross-section shape (i.e., concave versus convex), and distance to the deepest point, of approximately 5,000 waterbodies from bathymetric map data. We supplemented these geometry measures with data classifying waterbodies as reservoirs or natural lakes. We used this data to compute geometric depth estimates (Equation \ref{eq1}) and prediction "offsets" to these estimates using the random forest algorithm (Equation \ref{eq3}). Such offsets are model quantities which minimize differences between observed and predicted depth as a function of covariates. In our case, covariates included a variety of waterbody, watershed, and hydrologic subbasin measures that are available for all waterbodies (Table 1).

@@

## Methods
### Data description
We compiled bathymetry data on approximately 5,000 waterbodies in the Northeastern and Midwestern United States from nine official state databases (Fig. \ref{fig2}). These represent approximately 15% of all waterbodies in the States included in our study and are a diverse cross section in terms of their characteristics, surface areas, and span a wide geographic extent including glaciated and non-glaciated regions (Table 1). Thus, they can be considered representative of the entire population of lakes in our study extent.

@@

Covariates used in random forest modeling (Table 1, Equation 3, see Random Forests Model sub-section below) for waterbody elevation, area, island area, perimeter, shoreline development, watershed to waterbody area ratio, and hydrologic subbasin (i.e. HUC4s), were obtained from the LAGOS-US LOCUS database. One such measure, that of shoreline development, is a measure of waterbody perimeter shape defined as:

@@

### Proxy evaluation
We conducted a qualitative assessment of whether or not proxy measures of waterbody geometry (e.g.  nearshore land slope, distance to the center of the waterbody) are representative of their true values (e.g. in-lake slope, distance to the deepest point of the waterbody) by visual inspection (i.e. plotting each proxy measure against its corresponding true value) and by computing coefficients of determination ($R^2$). We further tested proxy measures by examining their effect on waterbody depth prediction error. Our approach involved several steps. In the first step, we computed a geometric estimate of waterbody depth using only geometry information ($depth_{geometric}$, Equation \ref{eq1}). In the second step, we fit a random forest model to predict observed (i.e. true) depth as a function of geometric depth along with several covariates available for all waterbodies (Table 1). The purpose of this random forest “offset” modeling was to more rigorously test our expectations regarding prediction error among different formulations of $depth_{geometric}$ and among different waterbody types. Each of these steps were executed iteratively for each combination of true and proxy values of slope and distance (Table 2). We conducted additional sensitivity analysis to examine possible interactions between different proxy measures of waterbody geometry and different subsets of the entire dataset where model data was restricted (i.e. “filtered”) to include only reservoirs, only natural lakes, only convex waterbodies, or only concave waterbodies (Table 2).

@@

Our tests of the geometric cone model of waterbody depth models show that specific proxy measures of waterbody geometry are not representative of true geometry measures across a broad array of waterbodies. Models using non-representative proxies showed increased error and systematic overestimation of depth in concave and reservoir waterbodies. Although our analysis was limited to waterbodies with available bathymetry data, these waterbodies did not have characteristics that differed from that of the overall waterbody population (apart from the fact that our study waterbodies were somewhat larger in area compared to the overall waterbody population, See Fig. S7, Table 1, Fig. S10). Although there is a possibility that there is some hidden bias not explored for in our analyses, this lack of difference suggests that our results are likely to be broadly applicable to nearly all waterbodies in the study area.

>>>>

As a result of this limited testing, we lack knowledge on both the predictive performance of geometric models, the effect of proxies on depth prediction, and whether depth predictions are more sensitive to measurement errors in the horizontal dimension (i.e. distance to the deepest point of the waterbody) or measurement errors in the vertical dimension (i.e. in-lake slope). Additionally, it is unclear whether model prediction error is related to differences in waterbody type such those with different cross-section shapes (concave versus convex) or those classified as reservoirs versus natural lakes. To address these knowledge gaps, we asked three research questions: (1) How representative is nearshore land slope of in-lake slope; and how representative is the distance to the center of a waterbody compared to the distance to the deepest point of a waterbody? (2) How does the use of proxies for waterbody geometry affect waterbody depth prediction error? (3) How does waterbody cross-section shape (i.e. concave versus convex) and waterbody type (i.e. natural lake vs reservoir) affect depth prediction error? To answer these questions, we extracted maximum depth (hereafter referred to as “observed maximum depth”), in-lake slope, cross-section shape (i.e., concave versus convex), and distance to the deepest point, of approximately 5,000 waterbodies from bathymetric map data. We supplemented these geometry measures with data classifying waterbodies as reservoirs or natural lakes. We used this data to compute geometric depth estimates (Equation \ref{eq1}) and prediction "offsets" to these estimates using the random forest algorithm (Equation \ref{eq3}). Such offsets are model quantities which minimize differences between observed and predicted depth as a function of covariates. In our case, covariates included a variety of waterbody, watershed, and hydrologic subbasin measures that are available for all waterbodies (Table \ref{table1}).

\begin{table}
  \caption{Summary of lake characteristics for the present study (and for lakes in the contiguous United States). Predictor variables for computing random forest offsets (Equation \ref{eq2}) are printed in bold face. Dashes (-) indicate an identical sample size among this study and that of the contiguous United States from the National Hydrography Dataset. The total number lakes is reported as n.} \label{table1}
  \centering
  \setlength\tabcolsep{1.5pt} % default value: 6pt
  \begin{tabular}{lllll}
    \hline
    Variable & Median & Q25 & Q75 & n\\
    \hline
    Max depth (m) & 8.2 (7) & 4.6 (3.7) & 14 (12) & 4820 (17700)\\
    \textbf{Area (ha)} & 55 (33) & 21 (11) & 140 (100) & 4820 (17700)\\
    \textbf{Island area (ha)} & 0 (0) & 0 (0) & 0.18 (0.076) & 4820 (17700)\\
    \textbf{Perimeter (m)} & 4400 (3500) & 2500 (1800) & 8100 (7300) & 4820 (17700)\\
    \textbf{Shoreline development} & 1.7 (1.7) & 1.4 (1.4) & 2.1 (2.2) & 4820 (17700)\\
    \textbf{Elevation (m)} & 300 (340) & 180 (210) & 400 (460) & 4820 (17700)\\    
    \textbf{Watershed-lake ratio} & 7.8 (10) & 3.9 (4.4) & 17 (28) & 4820 (17700)\\
    Deepest point distance (m) & 180 (-) & 110 (-) & 290 (-) & 4820 (-)\\
    Mean deepest point distance (m) & 140 (-) & 87 (-) & 230 (-) & 4820 (-)\\
    \textbf{Visual center distance(m)} & 240 (-) & 160 (-) & 390 (-) & 4820 (-)\\
    Inlake slope (m/m) & 0.05 (-) & 0.02 (-) & 0.08 (-) & 4820 (-)\\
    Inlake slope online (m/m) & 0.06 (-) & 0.03 (-) & 0.14 (-) & 4800 (-)\\
    Inlake slopes (m/m) & 0.06 (-) & 0.03 (-) & 0.1 (-) & 4820 (-)\\
    Inlake slopes online (m/m) & 0.07 (-) & 0.03 (-) & 0.15 (-) & 4800 (-)\\
    Mean inlake slope (m/m) & 0.04 (-) & 0.02 (-) & 0.09 (-) & 4820 (-)\\
    Nearshore mean slope (m/m) & 0.08 (-) & 0.05 (-) & 0.11 (-) & 4820 (-)\\
    Nearshore slope online (m/m) & 0.08 (-) & 0.04 (-) & 0.13 (-) & 4590 (-)\\
    Nearshore slopes online (m/m) & 0.08 (-) & 0.04 (-) & 0.13 (-) & 4540 (-)\\
    \bottomrule
    \end{tabular}  
  \end{table}

@@

## Methods
### Data description
We compiled bathymetry data on approximately 5,000 waterbodies in the Northeastern and Midwestern United States from nine official state databases (Fig. \ref{fig2}). These represent approximately 15% of all waterbodies in the States included in our study and are a diverse cross section in terms of their characteristics, surface areas, and span a wide geographic extent including glaciated and non-glaciated regions (Table \ref{table1}). Thus, they can be considered representative of the entire population of lakes in our study extent.

@@

Covariates used in random forest modeling (Table \ref{table1}, Equation 3, see Random Forests Model sub-section below) for waterbody elevation, area, island area, perimeter, shoreline development, watershed to waterbody area ratio, and hydrologic subbasin (i.e. HUC4s), were obtained from the LAGOS-US LOCUS database. One such measure, that of shoreline development, is a measure of waterbody perimeter shape defined as:

@@

### Proxy evaluation
We conducted a qualitative assessment of whether or not proxy measures of waterbody geometry (e.g.  nearshore land slope, distance to the center of the waterbody) are representative of their true values (e.g. in-lake slope, distance to the deepest point of the waterbody) by visual inspection (i.e. plotting each proxy measure against its corresponding true value) and by computing coefficients of determination ($R^2$). We further tested proxy measures by examining their effect on waterbody depth prediction error. Our approach involved several steps. In the first step, we computed a geometric estimate of waterbody depth using only geometry information ($depth_{geometric}$, Equation \ref{eq1}). In the second step, we fit a random forest model to predict observed (i.e. true) depth as a function of geometric depth along with several covariates available for all waterbodies (Table \ref{table1}). The purpose of this random forest “offset” modeling was to more rigorously test our expectations regarding prediction error among different formulations of $depth_{geometric}$ and among different waterbody types. Each of these steps were executed iteratively for each combination of true and proxy values of slope and distance (Table 2). We conducted additional sensitivity analysis to examine possible interactions between different proxy measures of waterbody geometry and different subsets of the entire dataset where model data was restricted (i.e. “filtered”) to include only reservoirs, only natural lakes, only convex waterbodies, or only concave waterbodies (Table 2).

\begin{table}
  \caption{Model fit and predictive accuracy metrics (RMSE = root mean square error, R2 = coefficient of determination, MAPE = mean absolute percent error) for all combinations of true (in-lake slope, distance to the deepest point of the waterbody) and proxy (nearshore land slope, distance to waterbody center) metrics.} \label{table2}
  \centering
  % \setlength\tabcolsep{1.5pt} % default value: 6pt
\begin{tabular}{llll}
  \hline
  slope & distance & RMSE & $R^2$\\
  \hline
  true & true & - & -\\
  true & proxy & 4.2 m & 0.73\\
  proxy & true & 6.9 m & 0.26\\
  proxy & proxy & 6.6 m & 0.31\\
  \bottomrule
\end{tabular}
\end{table}

@@

Our tests of the geometric cone model of waterbody depth models show that specific proxy measures of waterbody geometry are not representative of true geometry measures across a broad array of waterbodies. Models using non-representative proxies showed increased error and systematic overestimation of depth in concave and reservoir waterbodies. Although our analysis was limited to waterbodies with available bathymetry data, these waterbodies did not have characteristics that differed from that of the overall waterbody population (apart from the fact that our study waterbodies were somewhat larger in area compared to the overall waterbody population, See Fig. S7, Table \ref{table1}, Fig. S10). Although there is a possibility that there is some hidden bias not explored for in our analyses, this lack of difference suggests that our results are likely to be broadly applicable to nearly all waterbodies in the study area.
